Diffusion local times and the time until recombination

### 2015-10-22 notes/diffusion-local-times.Rmd

*Useful functions used below:*
```{r setup_fns}
pcline <- function (x,r,tau=1,sigma=1) {
    (1/2)*( 1 - sign(x)*zeta(x=x/sqrt(tau*sigma^2),r=r*tau) )
}
tanh_cline <- function (x,s,sigma=1) {
    (1/2)*(1+tanh(-2*x*sqrt(s)/sigma))
}
tanh_drift <- function (x,s,sigma=1) {
    # log-derivative of tanh_cline
    -(2*sqrt(s)/sigma) * (1-tanh(-2*x*sqrt(s)/sigma)^2) / (1+tanh(-2*x*sqrt(s)/sigma))
}
tanh_het <- function (x,s,sigma=1) {
    # prob of being in a het = (1-s)(1-p)/(1-s(1-p))
    pp <- tanh_cline(x,s,sigma=sigma)
    (1-s)*(1-pp)/(1-s*(1-pp))
}
sim_diffusion <- function (drift,tau,nwalks,ntimes,xinit=0,...) {
    # want total time to be tau, so:
    dt <- tau/ntimes
    # put the noise in rws
    rws <- matrix( c( rep_len(xinit,nwalks), # initial conditions
                 rnorm(nwalks*(ntimes-1)) ), nrow=ntimes, ncol=nwalks, byrow=TRUE )
    # and then integrate over it
    for (k in 2:ntimes) {
        rws[k,] <- rws[k-1,]+sqrt(dt)*rws[k,]+dt*drift(rws[k-1,],...)
    }
    attr(rws,"dt") <- dt
    return(rws)
}
weighted_time <- function (rws, weight.fn, use.t=TRUE, ...) {
    colSums( weight.fn(rws[use.t,],...) ) * attr(rws,"dt")
}
```

@@

The key quantity we compute with diffusion theory 
is the probability that a lineage, 
begun at location $x$ and killed at rate $r$ when in heterozygotes,
is still alive at time $\tau$.
Denote this by
$$
\begin{aligned}
    \zeta(x,r,\tau,\sigma).
\end{aligned}
$$

Now, let $p(x,r,\tau,A)$ denote the probability that a lineage begun at $x<0$
at a location at recombination distance $r$ from the selected site
inherits from the left-hand side of the barrier
(i.e., an ancestor on $(-\infty,0)$)
$\tau$ generations ago.
(Note that $r$ should be scaled by the density and fecundity of heterozygotes in the cline --
see below.)
For an individual to the left of the barrier ($x<0$),
the probability of inheriting from the left of the barrier at $\tau$
is the probability that no recombination at the barrier has occurred,
plus one-half the probability that recombination in a heterozygote has occurred,
i.e., $\zeta +(1-\zeta)/2=(1+\zeta)/2$.
If $x>0$ this is the one-half times probability a recombination in a heterozygote has occurred,
i.e., $(1-\zeta)/2$.
Combining these,
the probability of ending up on the left-hand side is
$$
\begin{aligned}
    p(x,r,\tau,A) = 
    \frac{1}{2} \left\{ 1 - \sgn(x) \zeta(x,r,\tau) \right\} .
\end{aligned}
$$

@@

Now, to solve the real problem:
```{r solve_pde, cache=TRUE, depends="solve_pde_setup"}
pde.fn <- function (t,y,parms,...) {
    tran <- tran.1D(C=y, A=clinefn, D=1/2, dx=thegrid)$dC
    list( tran - r * hetfn$mid * y )
}
pde.soln <- ode.1D( y=yinit, times=tt, func=pde.fn, nspec=1 )
image(pde.soln,legend=TRUE,main="prob of no recomb by t", yaxt='n', ylab='space')
axis(2,at=(pretty(thegrid$x.mid)-min(thegrid$x.mid))/diff(range(thegrid$x.mid)),labels=pretty(thegrid$x.mid))
```

@@

@@

Let's check that, too: why not?
```{r killing, cache=TRUE, depends="sim_rw"}
loc <- 0.1
zeta <- function (x,r) {
    2*pnorm(abs(x)) - 1 + 2 * exp( r*abs(x) + r^2/2 + pnorm(abs(x)+r,lower.tail=FALSE,log.p=TRUE) )
}
compare.rates <- as.data.frame( t( sapply( c(0.25,0.5,1,1.5,2,2.5,3), function (rate) {
        visits <- colSums(abs(rws-loc)<=1.5*eps)
        killed <- ( rbinom(ncol(rws),size=visits,prob=rate/ntimes/(3*eps)) == 0 )
        c( 
          rate=rate,
          simulation=mean(killed), 
          theory=zeta(loc,rate),
          SE=sqrt(mean(killed)*(1-mean(killed))/nwalks)
          )
    } ) ) )
with(compare.rates, {
         plot( rate, theory, type='l', xlab="recomb rate", ylab="prob of recomb", col='red', ylim=c(0,1) ) 
         points( rate, simulation, pch=20 )
         arrows( x0=rate, y0=simulation-2*SE, y1=simulation+2*SE, angle=90, code=3 )
    } )
```

@@

More generally, here's what $\zeta$ looks like:
```{r plot_zeta, fig.height=2*fig.dim}
layout(1:2)
xvals <- seq(0,5,length.out=150)
rvals <- seq(0,10,length.out=100)
zetavals <- outer( xvals, rvals, zeta )
matplot(xvals,zetavals,type='l',xlab="x",ylab=expression(zeta),col=rainbow(length(rvals)))
lrvals <- seq(1,length(rvals),length.out=6)
legend("bottomright",lty=1,col=rainbow(length(rvals))[lrvals],legend=paste("r=",round(rvals[lrvals],digits=2)))
matplot(rvals,t(zetavals),type='l',xlab="r",ylab=expression(zeta),col=rainbow(length(xvals)))
lxvals <- seq(1,length(xvals),length.out=6)
legend("bottomright",lty=1,col=rainbow(length(xvals))[lxvals],legend=paste("x=",round(xvals[lxvals],digits=2)),bg='white')
```

Now, let $p(x,r,\tau,A)$ denote the probability that a lineage begun at $x<0$
at a location at recombination distance $r$ from the selected site
inherits from the left-hand side of the barrier
(i.e., an ancestor on $(-\infty,0)$)
$\tau$ generations ago.
(Note that $r$ should be scaled by the density and fecundity of heterozygotes in the cline --
see below.)
For an individual to the left of the barrier ($x<0$),
the probability of inheriting from the left of the barrier at $\tau$
is the probability that no recombination at the barrier has occurred,
plus one-half the probability that recombination in a heterozygote has occurred,
i.e., $\zeta +(1-\zeta)/2=(1+\zeta)/2$.
If $x>0$ this is the one-half times probability a recombination in a heterozygote has occurred,
i.e., $(1-\zeta)/2$.
Combining these,
the probability of ending up on the left-hand side is
$$
\begin{aligned}
    p(x,r,\tau,A) = 
    \frac{1}{2} \left\{ 1 - \sgn(x) \zeta(x,r,\tau) \right\} .
\end{aligned}
$$

Here's the cline shape:
```{r show_cline, cache=TRUE}
pcline
xx <- seq(-15,15,length.out=500)
for (tau in c(5,50)) {
    plot(xx, pcline(xx,r=1), type='n', 
            xlab="distance from cline center", ylab="prob inherited from the left", 
            main=as.expression(substitute(tau==mytau,list(mytau=tau))) )
    rr <- seq(0,0.2,length.out=11)
    pp <- sapply( rr, function (r) { pcline(xx,r=r,tau=tau) } )
    matplot(xx,pp,type='l',lty=1,col=rainbow(length(rr)),add=TRUE)
    legend("topright",lty=1,col=rainbow(length(rr)),legend=paste("r=",rr))
}
```

>>>>

*Useful functions used below:*
```{r setup_fns}
pcline <- function (x,r,recombfn,...) {
    (1/2)*( 1 - sign(x)*recombfn(x,2*r,...) )
}
tanh_cline <- function (x,s,sigma=1) {
    (1/2)*(1+tanh(-2*x*sqrt(s)/sigma))
}
tanh_drift <- function (x,s,sigma=1) {
    # log-derivative of tanh_cline
    -(2*sqrt(s)/sigma) * (1-tanh(-2*x*sqrt(s)/sigma)^2) / (1+tanh(-2*x*sqrt(s)/sigma))
}
tanh_het <- function (x,s,sigma=1) {
    # prob of being in a het = (1-s)(1-p)/(1-s(1-p))
    pp <- tanh_cline(x,s,sigma=sigma)
    (1-s)*(1-pp)/(1-s*(1-pp))
}
zeta.reflected <- function (x,r) {
    2*pnorm(abs(x)) - 1 + 2 * exp( r*abs(x) + r^2/2 + pnorm(abs(x)+r,lower.tail=FALSE,log.p=TRUE) )
}
sim_diffusion <- function (drift,tau,nwalks,ntimes,xinit=0,...) {
    # want total time to be tau, so:
    dt <- tau/ntimes
    # put the noise in rws
    rws <- matrix( c( rep_len(xinit,nwalks), # initial conditions
                 rnorm(nwalks*(ntimes-1)) ), nrow=ntimes, ncol=nwalks, byrow=TRUE )
    # and then integrate over it
    for (k in 2:ntimes) {
        rws[k,] <- rws[k-1,]+sqrt(dt)*rws[k,]+dt*drift(rws[k-1,],...)
    }
    attr(rws,"dt") <- dt
    return(rws)
}
weighted_time <- function (rws, weight.fn, use.t=TRUE, ...) {
    colSums( weight.fn(rws[use.t,],...) ) * attr(rws,"dt")
}
clineplot <- function (soln,x=thegrid$x.mid,times=seq(1,nrow(soln),length.out=ntimes),ntimes=20,ylab="probability",ylim=c(0,1),...) {
    # plot output of deSolve as clines in a nice way
    graphics::matplot( x, t(soln[times,-1]), type='l', col=rainbow(length(times)), lty=1, 
            xlab="space", ylab=ylab, ylim=ylim, ...)
    legend("bottomleft", lty=1, col=rainbow(length(times)), legend=paste("t=",round(soln[times,1],digits=2)), cex=0.5)
    return( invisible( t(soln[times,-1]) ) )
}
```

@@

The key quantity we compute with diffusion theory 
is the probability that a lineage, 
begun at location $x$ and killed at rate $k$ when in heterozygotes,
is still alive at time $\tau$.
Denote this by
$$
\begin{aligned}
    \zeta(x,k,\tau).
\end{aligned}
$$
Now, let $q(x,r,\tau,A)$ denote the probability that a lineage begun at $x<0$
at a location at recombination distance $r$ from the selected site
inherits from the left-hand side of the barrier
(i.e., an ancestor on $(-\infty,0)$)
$\tau$ generations ago.
The probability that a recombination occurs 
between the lineage and the selected locus on either pair of chromatids
is $2r$,
and if this happens, the lineage ends up on either background with probability $1/2$.
Therefore,
for an individual to the left of the barrier ($x<0$),
the probability of inheriting from the left of the barrier at $\tau$
is the probability that no recombination at the barrier has occurred,
plus one-half the probability that recombination in a heterozygote has occurred,
so if $\zeta = \zeta(x,2r,\tau)$, then
$q=\zeta +(1-\zeta)/2=(1+\zeta)/2$.
If $x>0$ this is the one-half times probability a recombination in a heterozygote has occurred,
i.e., $q=(1-\zeta)/2$.
Combining these,
the probability of ending up on the left-hand side is
$$
\begin{aligned}
    q(x,r,\tau,A) = 
    \frac{1}{2} \left\{ 1 - \sgn(x) \zeta(x,2r,\tau) \right\} .
\end{aligned}
$$

@@

Now, to solve the real problem:
```{r solve_pde, cache=TRUE, depends="solve_pde_setup"}
pde.fn <- function (t,y,parms,...) {
    tran <- tran.1D(C=y, A=clinefn, D=1/2, dx=thegrid)$dC
    list( tran - r * hetfn$mid * y )
}
pde.soln <- ode.1D( y=yinit, times=tt, func=pde.fn, nspec=1 ) # note FIRST COLUMN IS TIME
image(pde.soln,legend=TRUE,main="prob of no recomb by t", yaxt='n', ylab='space')
axis(2,at=(pretty(thegrid$x.mid)-min(thegrid$x.mid))/diff(range(thegrid$x.mid)),labels=pretty(thegrid$x.mid))
```

@@

The even more right solution
============================

Actually, the cline gets set up
in a time-inhomogeneous way.
Let's do that, by numerical solution,
then run the lineage back through it.
The forwards-time evolution of the cline,
with isotropic density and migration,
solves the PDE (Barton 1979):
$ \begin{align*}
  \partial_t p(x,t) = \frac{\sigma^2}{2} \partial_x^2 + sp(x,t)(1-p(x,t))(2p(x,t)-1) .
\end{align*} $
```{r fwds_pde, cache=TRUE}
thegrid <- setup.grid.1D(x.up=-15, x.down=15, N = 5000)
yinit <- ifelse( thegrid$x.mid>0, 0, 1 )
tt <- seq(0,tau,length.out=100)
fwds.pde <- function (t,y,parms,...) {
    tran <- tran.1D(C=y, D=1/2, dx=thegrid)$dC
    list( tran + s * y * (1-y) * (2*y-1) )
}
fwds.soln <- ode.1D( y=yinit, times=tt, func=fwds.pde, nspec=1 ) # note FIRST COLUMN IS TIME
image(fwds.soln,legend=TRUE,main="mean cline development", yaxt='n', ylab='space')
axis(2,at=(pretty(thegrid$x.mid)-min(thegrid$x.mid))/diff(range(thegrid$x.mid)),labels=pretty(thegrid$x.mid))
```
Now, let's run the killed lineage backwards in it:
```{r fwds_back_pde, cache=TRUE, depends="fwds_pde"}
yinit <- rep(1.0,thegrid$N)
cline.interp <- function (T) {
    # return the interpolated cline for a particular time
    which.t <- findInterval(T,fwds.soln[,1],rightmost.closed=TRUE)
    lu <- fwds.soln[c(which.t,which.t+1),1]
    alpha <- (T-lu[1])/diff(lu)
    return( (1-alpha)*fwds.soln[which.t,-1] + alpha*fwds.soln[which.t+1,-1]  )
}
rev.pde <- function (t,y,parms,...) {
    p <- cline.interp(tau-t)
    # p <- setup.prop.1D(grid=thegrid, xy=cbind(thegrid$x.mid,cline.interp(tau-t)), interp="linear")
    tran <- tran.1D(C=y, A=p, D=1/2, dx=thegrid)$dC
    list( tran - r * (1-p) * y )
}
rev.soln <- ode.1D( y=yinit, times=tt, func=rev.pde, nspec=1, tcrit=tau ) # note FIRST COLUMN IS TIME
image(rev.soln,legend=TRUE,main="prob of no recomb by t", yaxt='n', ylab='space')
axis(2,at=(pretty(thegrid$x.mid)-min(thegrid$x.mid))/diff(range(thegrid$x.mid)),labels=pretty(thegrid$x.mid))
```

Let's compare the case where the cline is set up instantly or not:
```{r compare_rightness, cache=TRUE,depends=c("fwds_back_pde","solve_pde_setup","setup_fns")}
layout(t(1:2))
fb.p <- clineplot(rev.soln,main="forwards-backwards")
st.p <- clineplot(pde.soln,main="stationary cline")
layout(1)
plot(fb.p,st.p,asp=1,pch=20,cex=0.5,xlab="forwards-backwards",ylab="stationary",col=rainbow(ncol(fb.p))[col(fb.p)])
abline(0,1,lwd=3)
```

@@

Let's check that, too: why not?
```{r killing, cache=TRUE, depends="sim_rw"}
zeta.reflected
loc <- 0.1
compare.rates <- as.data.frame( t( sapply( c(0.25,0.5,1,1.5,2,2.5,3), function (rate) {
        visits <- colSums(abs(rws-loc)<=1.5*eps)
        killed <- ( rbinom(ncol(rws),size=visits,prob=rate/ntimes/(3*eps)) == 0 )
        c( 
          rate=rate,
          simulation=mean(killed), 
          theory=zeta.reflected(loc,rate),
          SE=sqrt(mean(killed)*(1-mean(killed))/nwalks)
          )
    } ) ) )
with(compare.rates, {
         plot( rate, theory, type='l', xlab="recomb rate", ylab="prob of recomb", col='red', ylim=c(0,1) ) 
         points( rate, simulation, pch=20 )
         arrows( x0=rate, y0=simulation-2*SE, y1=simulation+2*SE, angle=90, code=3 )
    } )
```

@@

More generally, here's what $\zeta$ looks like:
```{r plot_zeta, fig.height=2*fig.dim}
layout(1:2)
xvals <- seq(0,5,length.out=150)
rvals <- seq(0,10,length.out=100)
zetavals <- outer( xvals, rvals, zeta.reflected )
matplot(xvals,zetavals,type='l',xlab="x",ylab=expression(zeta),col=rainbow(length(rvals)))
lrvals <- seq(1,length(rvals),length.out=6)
legend("bottomright",lty=1,col=rainbow(length(rvals))[lrvals],legend=paste("r=",round(rvals[lrvals],digits=2)))
matplot(rvals,t(zetavals),type='l',xlab="r",ylab=expression(zeta),col=rainbow(length(xvals)))
lxvals <- seq(1,length(xvals),length.out=6)
legend("bottomright",lty=1,col=rainbow(length(xvals))[lxvals],legend=paste("x=",round(xvals[lxvals],digits=2)),bg='white')
```

Recall from above that the predicted cline is
$$
\begin{aligned}
    q(x,r,\tau,A) = 
    \frac{1}{2} \left\{ 1 - \sgn(x) \zeta(x,2r,\tau) \right\} .
\end{aligned}
$$

Here's the cline shape:
```{r show_cline, cache=TRUE}
zeta.pcline <- function (x,r,recombfn=function(x,r){zeta.reflected(x=x/sqrt(tau*sigma^2),r=r*tau)},tau=1,sigma=1) {
    (1/2)*( 1 - sign(x)*zeta.reflected(x=x/sqrt(tau*sigma^2),r=r*tau) )
}
xx <- seq(-15,15,length.out=500)
for (tau in c(5,50)) {
    plot(xx, zeta.pcline(xx,r=1), type='n', 
            xlab="distance from cline center", ylab="prob inherited from the left", 
            main=as.expression(substitute(tau==mytau,list(mytau=tau))) )
    rr <- seq(0,0.2,length.out=11)
    pp <- sapply( rr, function (r) { zeta.pcline(xx,r=r,tau=tau) } )
    matplot(xx,pp,type='l',lty=1,col=rainbow(length(rr)),add=TRUE)
    legend("topright",lty=1,col=rainbow(length(rr)),legend=paste("r=",rr))
}
```
